---
title: "Waiting_for_a_psychologist_in_Region_Midtjylland"
author: "Emma-Marie Vitskov Jørgensen"
date: "28/03/2023 updated`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


## Purpose of the project
If you need a psychologist, the waiting times can be significantly long, which has been critiqued in the past years (source). The many weeks of waiting can cause the condition of the waiting patients to worsen, and this risk is not equally distributed through the municipalities, which makes the waiting time not only a problem for the individual patient, but also a problem of inequality. The purpose of this project is to investigate the waiting times to see, which municipalities has the longest waiting times, and how many people live in these "vulnerable" areas and can therefore risk to be affected by the long waits. 

## About the Data

I use two data sets:
  1) The first data set is called ```gadm36_DNK_2_sp_rds```and is a shape file from GADM. It is a spatial data frame containing information about the municipalities in Denmark including the municipality polygons. 
  2) The second data set is from Region Midt, and contains the waiting times in Region Midtjylland measured in weeks. The data is from a PDF found on xxxx, and I have manually typed the data into a excel file (saved as .csv), which can be found in the ```data```folder on Github. My pre-processing of the data is described in the project report.  

## Libraries

```{r library}
library(sf) 
library(raster)
library(dplyr)
library(readr)
library(mapview)
library(RColorBrewer)
library(cartogram)
library(tmap)

```


## Load data

Loading the municipalities shape file from GADM. I project the data to the EPSG number 25832 with the ```st_as_sf```:

```{r load-municipalities}
#load data
municipalities <-  getData("GADM", country = "DNK", level = 2)

class(municipalities) # data is a SpatialPolygons DataFrame

#project the data
municipalities_25832 <- municipalities %>% 
  st_as_sf(municipalities_25832) %>% 
  st_transform(municipalities_25832, crs = 25832) 

# inspecting the data
municipalities_25832
plot(municipalities_25832[7]) #plotting the municipalities geometry
class(municipalities_25832) # sf and data frame

# checking if the polygons are valid 
sf::st_is_valid(municipalities_25832) # they are all valid

```

Filtering the municipalities and only keeping the once in ```Midtjylland```. The regions are found in the ```NAME_1```column. Also changing the column name ```NAME_2``` to the more informative name municipalities: 

```{r Midtjylland-data}
#checking NAME_1 column containing the regions
municipalities_25832$NAME_1

#creating new data frame containing only data for Midtjylland
Midtjylland <- municipalities_25832[municipalities_25832$NAME_1 == 'Midtjylland',]

#change "NAME_2" to "municipality"
Midtjylland <- Midtjylland %>% 
        rename( "municipality" = "NAME_2")

sort(unique(Midtjylland$municipality))

# replace "Århus" with "Aarhus" to match the other datasets
Midtjylland$municipality[Midtjylland$municipality == "Århus"] <- "Aarhus"

```

## Waiting times in Region Midt
I will combine the municipalities data with my data of the waiting time in the different municipalities in Midtjylland. 

Loading the ```waitingtime_regionmidt.csv```data set and inspecting it. Changing the name "Aarhus" to "Århus to match the spelling of Århus in the ```Midtjylland```data: 

```{r load-waitingtime}
# load the waiting data
waitingtime <- read.csv2("../data/waitingtime_regionmidt.csv")

# Check names
sort(unique(waitingtime$municipality))

# fixing misspellings due to danish special characters
waitingtime$municipality[waitingtime$municipality == "Ringk\xbfbing-Skjern"] <- "Ringkøbing-Skjern"
waitingtime$municipality[waitingtime$municipality == "Sams\xbf"] <- "Samsø"

class(waitingtime) # it's a data frame

```

I will also add the number of people young people in the age of 18-24 in each municipality to the ```waiting-midt```dataset:

```{r load-population}
# load the waiting data

young_pop_midt <- read.csv2("../data/young_population_midt_2021.csv")

# change "Municipality" to "municipality"
young_pop_midt <- young_pop_midt %>% 
        rename( "municipality" = "Municipality")

# Check names
sort(unique(young_pop_midt$municipality))

#fixing misspellings due to danish special characters
young_pop_midt$municipality[young_pop_midt$municipality == "Ringk\xbfbing-Skjern"] <- "Ringkøbing-Skjern"
young_pop_midt$municipality[young_pop_midt$municipality == "Sams\xbf"] <- "Samsø"

young_pop_midt$total_youngpop = rowSums(young_pop_midt[,c(2,3,4)])

young_pop_midt

```

```{r load-practices-data}
practices <- read.csv2("../data/psycology_practices_2017.csv")

# Check names
sort(unique(practices$municipality))

#fixing misspellings due to danish special characters
practices$municipality[practices$municipality == "Ringk\xbfbing-Skjern"] <- "Ringkøbing-Skjern"
practices$municipality[practices$municipality == "Sams\xbf"] <- "Samsø"
```

## Merge data

Now I combine the four data sets, so that the municipalities in the four data sets are matched with each other:

```{r merge-dataframes}
# merge the two data frames
waiting_midt <- Midtjylland %>% 
  left_join(waitingtime) %>% 
  left_join(young_pop_midt) %>% 
  left_join(practices)

waiting_midt
View(waiting_midt)
```

Save the new merged data set in the ```data_output``` folder:
```{r save-merged-data}
setwd('..') #changing position
readr::write_csv(waiting_midt,"data_output/waiting_midt.csv") 

```

## Which municipalities has the longest waiting time?

Now that the data is ready, I will investigate the waiting time in each of the 19 municipalities in Midtjylland. I found, that the best way to do this is to visualize the waiting time with plot()

```{r plot-waiting}

mapview(waiting_midt[c(7, 14, 15, 16, 24)], zcol = "average_waiting", col.regions=rev(brewer.pal(10, "RdBu")))

```

## Cartogram

How many young people in the age of 18-24 lives in the different municipalities? Is there a correlation between long waiting time and a high number of young people in this age group living in the municipality?
How many people live in the "vulnerable" areas?

```{r cartogram}
 
# Visualize the young population (18-24 years) in each municipality
pop_normal_map <- tm_shape(waiting_midt["total_youngpop"]) + 
  tm_polygons("total_youngpop", palette = "YlGn", title = "Population age 18-24", style = "jenks") +
  tm_layout(frame = FALSE, legend.position = c("right", "top")) 

# construct a cartogram using the waiting times as the size and young population as color
waiting_cartogram <- cartogram(waiting_midt, "average_waiting", itermax=5)

pop_cartogram <- tm_shape(waiting_cartogram) + 
  tm_polygons("total_youngpop", palette = "YlGn", title = "Population age 18-24", style = "jenks") +
  tm_layout(frame = FALSE, legend.position = c("right", "top")) 

# Show the two maps
pop_normal_map
pop_cartogram


```

```{r save-maps-1}
# Save the two maps in the same pdf
pdf("../map_output/pop_normal_and_cartogram_maps")

tmap_arrange(pop_normal_map, pop_cartogram)

dev.off()
```

OR do this:

```{r}
# Visualize the young population (18-24 years) in each municipality
population_map <- tm_shape(waiting_midt["total_youngpop"]) + 
  tm_polygons("total_youngpop", palette = "YlGn", title = "Population age 18-24", style = "jenks") +
  tm_layout(frame = FALSE, legend.position = c("right", "top")) 

average_waiting_map <- tm_shape(waiting_midt["average_waiting"]) + 
  tm_polygons("average_waiting", palette = "Reds", title = "Witing time (weeks)", style = "jenks") +
  tm_layout(frame = FALSE, legend.position = c("right", "top")) 

# Show the two maps
population_map
average_waiting_map

```

Save the maps side by side:

```{r save-maps-2}
# Save the two maps in the same pdf
pdf("../map_output/waiting_and_population_maps")

tmap_arrange(population_map, average_waiting_map)

dev.off()
```

## Plot number of clinics

```{r practices-bubbles}

# create centroid for each municipality
centroids <- st_centroid(waiting_midt)

plot(waiting_midt$geometry);plot(centroids, add = TRUE)

# use tmap to create a bubble (connected to each centroids) with a size corresponding to the number of practices

tm_shape(waiting_cartogram) + 
  tm_polygons("total_youngpop", palette = "YlGn", title = "Population age 18-24", style = "jenks") +
  tm_layout(frame = FALSE, legend.position = c("right", "top")) +
  tm_bubbles(size = waiting_midt$average_waiting, col = "pink", alpha = 1, border.col = "red",
  border.lwd = 1, border.alpha = NA, scale = 1, size.lim = NA,
  sizes.legend = NULL, sizes.legend.labels = NULL, n = 5,
  style = "pretty", palette = NULL, labels = NULL,
  auto.palette.mapping = TRUE, contrast = 1, max.categories = 12)

```



```{r save bubble-map}

```



## Sources:
Merging data: https://www.statmethods.net/management/merging.html 

Rename column in data frame: https://sparkbyexamples.com/r-programming/rename-column-in-r/ 

Combine data frame with spatial data frame: https://stackoverflow.com/questions/56116443/how-do-i-combine-a-dataframe-with-a-spatial-dataframe-when-receiving-errors-with 

Create cartogram: https://cran.r-project.org/web/packages/cartogram/readme/README.html 

Color palette to mapview map: https://stackoverflow.com/questions/60099307/add-color-palette-to-mapview-map
